Utility of Diffusion of Innovations for Diagnosing the Spread of Conservation Behaviors

### 2022-03-12 DOI_Review/Manuscript.rmd
updating

date: "`r format(Sys.time(), '%B %d, %Y')`"
keywords: Community-based conservation, diffusion of innovations, cultural evolution, complex systems
bibliography: [./DOI_Conservation_Review.bib]
abstract: The study of community-based conservation is plagued by a large number of important variables, nonlinear dynamics, and interactions and feedbacks among predictors. This complexity has made the quantification of causal mechanisms and between-site comparative analyses notoriously difficult. Here, we argue that analyzing the emergence and persistence of community-based conservation institutions at the level of the individual, rather than the entire community, can yield important quantitative insights. We highlight the utility of Diffusion of Innovations theory (DOI) and the broader field of cultural evolution. In particular, the differential models of disease transmission, on which DOI is founded, may provide a productive avenue for quantifying the spread and persistence of community-based conservation institutions. We first simulate data on community adoption of a conservation institution, contingent on feedbacks between individual behavior and environmental processes. We then demonstrate that fitting these data to a probabilistic differential model of diffusion can produce reliable estimates of the rates of adoption, dropout, and long-term uptake from early on in the lifetime of an institution. Specifically, we show that the social learning mechanism individuals employ can alter the population pattern of adoption and the final proportion of individuals that adopt an intervention. Overall, we explore a new quantitative approach for modeling the spread of conservation behaviors using probabilistic differential equations and argue for the incorporation of cultural evolutionary theory into the field.

@@

```{r packages, include=FALSE, cache=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(rmarkdown)
library(bookdown)
library(formattable)
library(kableExtra)
library(dplyr)
library(magrittr)
library(prettydoc)
library(htmltools)
library(knitcitations)
library(devtools)
library(tidyverse)
library(scales)
library(extrafont)
```

@@

Many of the factors that support the long-term viability of conservation institutions are intertwined such that the applicability of one may depend on the level of another and some factors may be inextricably correlated [@agrawal_group_2001; @agrawal_sustainable_2003]. Predictors and outcomes of interest commonly exhibit threshold effects, time lags, feedbacks,  and a variety of other nonlinear dynamics, making causal inference through standard regression practices difficult [@folke_socialecological_2007; @levin_social-ecological_2013]. And further, which individual factors are most important is highly specific to a given community or location, suggesting that there is not possibly one unifying panacea for successful conservation institutions [@brooks_how_2012; @geist_proximate_2002; @agrawal_communities_2001]. This complexity of community-based resource management systems, as well as logistical constraints in measuring the sheer number of important variables has made multi-site comparative analyses elusive [@sesmad_social-ecological_2014; @biggs_routledge_2021]. Thus, even with decades of productive research on sustainable resource management, all major syntheses remain descriptive in nature, lacking a robust methodology to quantify causal insights [@frey_synthesis_2017; @mcginnis_social-ecological_2014; @gutierrez_leadership_2011].

We argue here that considering the stability of institutions within a population from the perspective of individuals, rather than entire communities may provide valuable insight for implementing community-based conservation projects, both qualitatively and quantitatively. Sustainable resource management institutions are observed at the community level, but in reality they are an emergent property of individual level decision making — communities do not adopt behaviors, individuals do [@currie_cultural_2021]. The normative conditions imposed on individuals by their ecological and social environments (i.e. the factors referenced above) affect the likelihood that they will endogenously engage in cooperative resource management practices or adopt them when exogenously introduced by an intervention.

@@

# Diffusion of innovations
Promoting the widespread adoption of some novel (to the individual) behavior is not an issue unique to community-based conservation; producers of cigarettes and snack foods, as well as public health professionals, have been grappling with these questions for quite some time. DOI recognizes that this process is primarily governed by the level of uncertainty an individual has  about the utility of the novel behavior and their personal threshold for the amount of certainty required to make a decision [@rogers_diffusion_2010]. When an individual is introduced to a new behavior, they often have a high degree of uncertainty as to the risks and benefits of adopting and will wait to adopt until that uncertainty is reduced to a tolerable level. The primary way individuals reduce uncertainty about the outcome of adopting a particular behavior is through second-hand observation of early adopters. Therefore, at the core of DOI is the observation that individuals imitate those around them when making decisions about adopting a novel behavior, rather than weighing the pros and cons independently [@henrich_search_2001].

@@

In the vast majority of empirical studies, the diffusion of a novel trait produces a sigmoidal “S” shape, where it initially spreads slowly, followed by a period of rapid adoption, then again spreads slowly until it reaches fixation [@rogers_diffusion_2010]. This “S” shaped curve is indicative of biased social learning, in contrast to the relatively rare “R” shaped curves observed when individuals learn through independent trial and error (fig 1) [@henrich_cultural_2001]. These two curve shapes however are not completely distinct. As the combination of learning strategies employed by individuals in a population shifts more toward individual or social learning, we expect to see a shift toward an “R” shape or “S” shape respectively [@henrich_cultural_2001].

@@

# Diffusion of innovations in community-based conservation
Just as the sigmoidal pattern indicative of social learning is seen in the majority of cases on behavioral adoption broadly, scholars have also noticed this pattern in over 80% of case studies on the uptake of community-based conservation interventions worldwide [@mascia_when_2018; @mills_how_2019]. Fittingly, high-level studies have begun identifying DOI as potentially useful for diagnosing why some community-based conservation interventions spread and persist, while others fizzle out. For example, in a review of 22 conservation interventions referenced above, the majority of initiatives were adopted by less than 30% of potential adopters at the sigmoidal equilibrium. The authors take insights from DOI to suggest that practitioners may actively spread information on interventions (as in Abernethy et al. [-@abernethy_two_2014]), or facilitate communication to increase adoption rates [@mills_how_2019; @greenhalgh_diffusion_2004]. Further, a recent framework for understanding the establishment, persistence, and spread of community‐based conservation practices identifies qualitative insights, such as intervention observability, from DOI as useful in guiding the development of conservation interventions [@mahajan_theorybased_2020].

@@

We use this model to simulate data on the adoption and persistence of a conservation institution given two different learning mechanisms, frequency-dependent and success biased. Aside from the learning mechanism employed by the population, all model conditions are identical for these two scenarios. Frequency-dependent transmission is used as a reference as it reliably produces the standard sigmoidal “S” curve of behavioral adoption and does not incorporate any feedbacks between adoption and the environment. Under the success-biased learning scenario, agents compare their harvests to others in each time step and disproportionately copy more successful agents. In this instantiation then, adoption is dependent on harvest success and harvest success is determined largely by the amount of available resources in each area. As the amount of available resources in an area is determined by the number of agents who harvest there, this learning mechanism incorporates behavior-environment feedbacks typical of natural resource management institutions. Because agents compare their own harvests to those of others in each time step under the success-biased learning scenario, rather than exclusively imitating other individuals, this scenario actually represents a mix of individual and social learning. The adoption pattern observed in this simulation results in a mix between an “R” and “S” type curve, reinforcing the interpretation of this scenario as a mix of individual and social learning [@henrich_cultural_2001]. Figure 3 shows the full output of the later simulation where adoption (enrollment) spikes initially, but drops off dramatically as the resources in the unconserved area are reduced following increased harvest pressure there as a result of the conservation intervention.

@@

![SIRS model fit of agent-based simulation data under unbiased social learning shown as translucent black lines representing 1000 draws from the posterior. Simulated data are overlaid with gray points signifying training data (first 30%) and blue points signifying test (later 70%) data.](./DirectCopying.png)

@@

@@

Coupled human and natural systems commonly exhibit dynamics with thresholds, resilience, and heterogeneity not present in social or natural systems in isolation and therefore are difficult to model analytically [@liu_complexity_2007]. In very complex cases such as these, we argue that agent-based simulations are the preferred approach as they allow for these features of complex systems which are difficult or even impossible to account for with process models such the SIRS  or SIRS extension model presented in equation 4 [@reyers_social-ecological_2018; @railsback_agent-based_2012; @wilensky_introduction_2015]. Agent-based simulations allow modelers to impose a realism in the governance or institutional dynamics that are largely intractable in closed-form mathematical equations [@bravo_agents_2011; joffre_combining_2015]. As the number of purely subsistence communities shrinks each year worldwide, and multiple levels of institutions are increasingly involved in natural resource management, incorporating these features into our models is paramount for reliable inference [@mckean_common_2000; @armitage_adaptive_2009].

>>>>

date: "`r format(Sys.time(), '%B %d, %Y')`"
keywords: Community-based conservation, diffusion of innovations, cultural evolution, complex systems
bibliography: [./DOI_Conservation_Review.bib]
abstract: The study of community-based conservation is plagued by a large number of important variables, nonlinear dynamics, and interactions and feedbacks among predictors. This complexity has made the quantification of causal mechanisms and between-site comparative analyses notoriously difficult. Here, we argue that analyzing the emergence and persistence of community-based conservation institutions at the level of the individual, rather than the entire community, can yield important quantitative insights. We highlight the utility of Diffusion of Innovations theory (DOI) and the broader field of cultural evolution. In particular, the differential models of disease transmission, on which DOI is founded, may provide a productive avenue for quantifying the spread and persistence of community-based conservation institutions. We first simulate data on community adoption of a conservation institution, contingent on feedbacks between individual behavior and environmental processes. We then demonstrate that fitting these data to a probabilistic differential model of diffusion can produce reliable estimates of the rates of adoption, dropout, and long-term uptake from early on in the lifetime of an institution. We emphasize that the social learning mechanism individuals employ can alter the population pattern of adoption and the final proportion of individuals that adopt an intervention. Overall, we explore a new quantitative approach for modeling the spread of conservation behaviors using probabilistic differential equations and argue for more complete incorporation of cultural evolutionary theory into the field.

@@

```{r packages, include=FALSE, cache=FALSE, message=FALSE, warning=FALSE}
#
library(knitr)
library(rmarkdown)
library(bookdown)
library(formattable)
library(kableExtra)
library(dplyr)
library(magrittr)
library(prettydoc)
library(htmltools)
library(knitcitations)
library(devtools)
library(tidyverse)
library(scales)
library(extrafont)
```

@@

Many of the factors that support the long-term viability of conservation institutions are intertwined such that the applicability of one may depend on the level of another and some factors may be inextricably correlated [@agrawal_group_2001; @agrawal_sustainable_2003]. Predictors and outcomes of interest commonly exhibit threshold effects, time lags, feedbacks,  and a variety of other nonlinear dynamics, making causal inference through standard regression practices difficult [@folke_socialecological_2007; @levin_social-ecological_2013]. And further, which individual factors are most important is highly specific to a given community or location, suggesting that there is not possibly one unifying panacea for successful conservation institutions [@brooks_how_2012; @geist_proximate_2002; @agrawal_communities_2001]. This complexity of community-based resource management systems, as well as logistical constraints in measuring the sheer number of important variables has made multi-site comparative analyses elusive [@sesmad_social-ecological_2014; @biggs_routledge_2021]. Thus, even with decades of productive research on sustainable resource management, all major syntheses remain descriptive in nature, lacking a robust methodology to quantify causal insights [@cumming_advancing_2020; @frey_synthesis_2017; @mcginnis_social-ecological_2014; @gutierrez_leadership_2011].

We argue here that considering the stability of institutions within a population from the perspective of individuals, rather than entire communities may provide valuable insight for implementing community-based conservation projects, both qualitatively and quantitatively. Sustainable resource management institutions are observed at the community level, but in reality they are an emergent property of individual level decision making — communities do not adopt behaviors, individuals do [@currie_cultural_2021]. The normative conditions imposed on individuals by their ecological and social environments (i.e. the factors referenced above) affect the likelihood that they will endogenously engage in cooperative resource management practices or adopt them when exogenously introduced by an intervention [@wilson_generalizing_2013; @andrews_cultural_2018].

@@

# Diffusion of innovations
Promoting the widespread adoption of some novel (to the individual) behavior is not an issue unique to community-based conservation; producers of cigarettes and snack foods, as well as public health professionals, have been grappling with these questions for quite some time. DOI recognizes that this process is primarily governed by the level of uncertainty an individual has  about the utility of the novel behavior and their personal threshold for the amount of certainty required to make a decision [@rogers_diffusion_2010]. When an individual is introduced to a new behavior, they often have a high degree of uncertainty as to the risks and benefits of adopting and will wait to adopt until that uncertainty is reduced to a tolerable level. The primary way individuals reduce uncertainty about the outcome of adopting a particular behavior is through second-hand observation of early adopters. Therefore, at the core of DOI is the observation that individuals imitate those around them when making decisions about adopting a novel behavior, rather than weighing the pros and cons independently [@rogers_diffusion_2010; @henrich_search_2001].

@@

In the vast majority of empirical studies, the diffusion of a novel trait produces a sigmoidal “S” shape, where it initially spreads slowly, followed by a period of rapid adoption, then again spreads slowly until it reaches fixation [@rogers_diffusion_2010]. This “S” shaped curve is indicative of biased social learning, in contrast to the relatively rare “R” shaped curves observed when individuals learn through independent trial and error (fig 1) [@henrich_cultural_2001]. The curvilinear “R” shape produced by independent, or “environmental” learning is a product of probability theory, in that late adopters represent the far tail of the distribution, having not discovered (and thus adopted) the novel behavior even after many trials [@henrich_cultural_2001]. These two curve shapes however are not completely distinct. As the combination of learning strategies employed by individuals in a population shifts more toward individual or social learning, we expect to see a shift toward an “R” shape or “S” shape respectively [@henrich_cultural_2001].

@@

# Diffusion of innovations in community-based conservation
Just as the sigmoidal pattern indicative of social learning is seen in the majority of cases on behavioral adoption broadly, scholars have also noticed this pattern in over 80% of case studies on the uptake of community-based conservation interventions worldwide [@mascia_when_2018; @mills_how_2019]. Fittingly, high-level studies have begun identifying DOI as potentially useful for diagnosing why some community-based conservation interventions spread and persist, while others fizzle out. For example, in a review of 22 conservation interventions referenced above, the majority of initiatives were adopted by less than 30% of potential adopters at the sigmoidal equilibrium. The authors take insights from DOI to suggest that practitioners may actively spread information on interventions [as in @abernethy_two_2014], or facilitate communication to increase adoption rates [@mills_how_2019; @greenhalgh_diffusion_2004]. Further, a recent framework for understanding the establishment, persistence, and spread of community‐based conservation practices identifies qualitative insights, such as intervention observability, from DOI as useful in guiding the development of conservation interventions [@mahajan_theorybased_2020].

@@

We use this model to simulate data on the adoption and persistence of a conservation institution given two different learning mechanisms, frequency-dependent and success-biased. Aside from the learning mechanism employed by the population, all model conditions are identical for these two scenarios. Frequency-dependent transmission is used as a reference as it reliably produces the standard sigmoidal “S” curve of behavioral adoption and does not incorporate any feedbacks between adoption and the environment. Under the success-biased learning scenario, agents compare their harvests to others in each time step and disproportionately copy more successful agents. In this instantiation then, adoption is dependent on harvest success and harvest success is determined largely by the amount of available resources in each area. As the amount of available resources in an area is determined by the number of agents who harvest there, this learning mechanism incorporates behavior-environment feedbacks typical of natural resource management institutions. Having independently sampled from one or both environments, agents then compare their harvests to those of others in each time step. In this way, the success-biased learning scenario actually represents a mix of individual and social learning, rather than a scenario where individuals make decisions purely from social information. The adoption pattern observed in this simulation results in a mix between an “R” and “S” type curve, reinforcing the interpretation of this scenario as a mix of individual and social learning [@henrich_cultural_2001]. Figure 3 shows the full output of the later simulation where adoption (enrollment) spikes initially, but drops off dramatically as the resources in the unconserved area are reduced following increased harvest pressure there as a result of the conservation intervention.

@@

![SIRS model fit of agent-based simulation data under frequency-dependent social learning shown as translucent black lines representing 1000 draws from the posterior. Simulated data are overlaid with gray points signifying training data (first 30%) and blue points signifying test (later 70%) data.](./DirectCopying.png)

@@

Lastly, while we have shown here that DOI can inform community-based conservation efforts in silico, the real-world may constrain such productive application. In a sharp response to Mascia and Mills [-@mascia_when_2018], authors Lund and Blewstein [-@lund_when_2018] question the narrative of “conservation interventions as innovations that can be freely adopted.” They argue that top-down support and manipulation are common in these contexts and may hinder the applicability of “demand-driven” conceptualizations of how conservation practices spread. These criticisms are paralleled by issues raised by Rogers himself when identifying the pro-innovation and individual blame biases of DOI, which portray lack of adoption as a flaw of the individual, opposed to the innovation [@rogers_diffusion_2010]. Given the problematic history of colonial-type conservation, we urge readers to seriously consider who is providing the narrative when assessing how and why conservation institutions diffuse within populations [@agrawal_enchantment_1999-1].

@@

Coupled human and natural systems commonly exhibit dynamics with thresholds, resilience, and heterogeneity not present in social or natural systems in isolation and therefore are difficult to model analytically [@liu_complexity_2007]. In very complex cases such as these, we argue that agent-based simulations are the preferred approach as they allow for these features of complex systems which are difficult or even impossible to account for with process models such the SIRS  or SIRS extension model presented in equation 4 [@reyers_social-ecological_2018; @railsback_agent-based_2012; @wilensky_introduction_2015]. Agent-based simulations allow modelers to impose a realism in the governance or institutional dynamics that are largely intractable in closed-form mathematical equations [@bravo_agents_2011; @joffre_combining_2015]. As the number of purely subsistence communities shrinks each year worldwide, and multiple levels of institutions are increasingly involved in natural resource management, incorporating these features into our models is paramount for reliable inference [@mckean_common_2000; @armitage_adaptive_2009].
